---
output:
  html_document:
    toc: true                               # table of contents
    toc_float: true                         # float the table of contents to the left of the main document content
    toc_depth: 3                            # header levels 1,2,3
    theme: default
    number_sections: false                  # add section numbering to headers
    df_print: paged                         # tables are printed as an html table with support for pagination over rows and columns
    css: ./nf-core_style.css
    highlight: pygments
    pdf_document: true
  html_notebook:
    toc: yes
---


---
title: "Repertoire analysis"
subtitle: "Interactive summary report produced by the nf-core/airrflow pipeline"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---


```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(tidyr)
library(alakazam)
library(shazam)
library(stringr)
library(plotly)

theme_set(theme_bw(base_family = "ArialMT") +
            theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), text = element_text(family="ArialMT")))
knitr::opts_chunk$set(echo = FALSE)

# Setting data table and creating output folders
datadir <- "."
outdir <- "repertoire_comparison"
seq_dir <- paste(outdir, "Sequence_numbers_summary", sep="/")
dir.create(seq_dir)

```
***

# Number of sequences

## Sequence assembly steps

Number of reads for each of the samples and number of sequences left after performing sequence assembly and alignment to reference data.
The full table can be found under [Table_sequences_assembly]("./Sequence_numbers_summary/Table_sequences_assembly.tsv").

```{r seq_numbers, echo=FALSE, warning=FALSE, results='asis'}
read_table <- function(tab_file){
        tab_seqs <- read.table(tab_file, header=TRUE, sep="\t", check.names = FALSE)
        write.table(tab_seqs, file=paste0(seq_dir,"/Table_sequences_assembly.tsv"), sep="\t", quote=F, row.names=F)
    }
tryCatch( {read_table("./Table_sequences.tsv")} ,
  error=function(e){message("No sequence numbers are available if starting with assembled reads.")}
)

```


```{r seq_numbers_plot, echo=FALSE, warning=FALSE, results='asis'}
tryCatch( {
    tab_seqs <- read.table("./Table_sequences.tsv", header=TRUE, sep="\t", check.names = FALSE)

    plot_table <- tidyr::pivot_longer(tab_seqs,
                                      cols=Sequences_R1:Igblast,
                                      names_to = "steps",
                                      values_to ="count")
    plot_table$steps <- factor(plot_table$steps, levels=c("Sequences_R1", "Sequences_R2", "Filtered_quality_R1",
                                                            "Filtered_quality_R2", "Mask_primers_R1", "Mask_primers_R2",
                                                            "Paired", "Build_consensus", "Assemble_pairs", "Unique",
                                                            "Representative_2", "Igblast"))
    seqs_plot <- ggplot(data=plot_table,
                        aes(x=steps, y=count, group=sample_id)) +
                  geom_line(aes(colour=sample_id)) +
                  geom_point(aes(colour=sample_id)) +
                  xlab("") + ylab("Sequence count") +
                  ggtitle("Number of sequences after each of the sequence assembly steps") +
                  theme(axis.text.x= element_text(angle = 45))

    ggplotly(seqs_plot)
  },
  error=function(e){message("No sequence numbers are available if starting with assembled reads.")}
)
```

## Filtering and clonal analysis steps

Number of sequences for each of the samples after each of the downstream filtering and clonal analysis steps.
The full table can be found under [Table_sequences_assembled]("./Sequence_numbers_summary/Table_sequences_assembled.tsv").

```{r seq_assembled, echo=FALSE, warning=FALSE, results='asis'}
tab_seqs_assembled <- read.table("./Table_sequences_assembled.tsv", header=TRUE, sep="\t", check.names = FALSE)

if (any(is.na(tab_seqs_assembled$sample_id))) {
  tab_seqs_assembled$sample_id <- sapply(tab_seqs_assembled$file_0, function(x) unlist(strsplit(as.character(x), "_"))[1])
}

dat <- tidyr::pivot_wider(tab_seqs_assembled,
                          id_cols=sample_id,
                          names_from=task,
                          values_from=to_num_seqs)
dat <- dat %>% dplyr::relocate(any_of(c("sample_id","ConvertDb-fasta", "AssignGenes-igblast", "MakeDB-igblast", "FilterQuality",
                                "ParseDb-split", "FilterJunctionMod3","AddMetadata","SingleCellQC","CreateGermlines",
                                "RemoveChimeric","CollapseDuplicates","ClonePass"))) %>% dplyr::arrange(sample_id)


write.table(dat, file=paste0(seq_dir,"/Table_sequences_assembled.tsv"), sep="\t", quote=F, row.names=F)
```


```{r assembled_seq_numbers_plot, echo=FALSE, warning=FALSE, results='asis'}

tab_seqs_assembled$task <- factor(tab_seqs_assembled$task, levels=c("AssignGenes-igblast", "MakeDB-igblast",
                                                                    "FilterQuality",
                                                          "ParseDb-split", "FilterJunctionMod3", "AddMetadata",
                                                          "CreateGermlines", "RemoveChimeric", "CollapseDuplicates",
                                                          "ClonePass"))
tab_seqs_assembled <- tab_seqs_assembled[!grepl("productive-F",tab_seqs_assembled$to_name),]

seqs_plot_assembled <- ggplot(data=tab_seqs_assembled,
                    aes(x=task, y=to_num_seqs, group=sample_id)) +
              geom_line(aes(colour=sample_id)) +
              geom_point(aes(colour=sample_id)) +
              xlab("") + ylab("Sequence count") +
              ggtitle("Number of sequences after each of the downstream steps") +
              theme(axis.text.x= element_text(angle = 45))

ggplotly(seqs_plot_assembled)
```

```{r read_data, include=FALSE}
### Read all the tables as produced by the pipeline in the current folder and joins them together in the df_all dataframe

all_files <- system(paste0("find '",datadir,"' -name '*clone-pass.tsv'"), intern=T)

dir.create(outdir)
diversity_dir <- paste(outdir, "Diversity", sep="/")
abundance_dir <- paste(outdir, "Abundance", sep="/")
vfamily_dir <- paste(outdir, "V_family", sep="/")
dir.create(diversity_dir)
dir.create(abundance_dir)
dir.create(vfamily_dir)

# Generate one big dataframe from all patient dataframes

df_list = lapply(all_files, read.csv, sep="\t")

df_all <- dplyr::bind_rows(df_list)

# Remove underscores in these columns
df_all$subject_id <- sapply(df_all$subject_id, function(x) str_replace(as.character(x), "_", ""))
df_all$sample_id <- sapply(df_all$sample_id, function(x) str_replace(as.character(x), "_", ""))

# Annotate sample and samplepop (sample + population) by add ing all the conditions
df_all$subj_locus <- as.factor(paste(df_all$sample_id, df_all$subject_id, df_all$pcr_target_locus, sep="_"))

# Write table to file
write.table(df_all, paste0(outdir,"/all_data.tsv"), sep = "\t", quote=F, row.names = F, col.names = T)

# Set number of bootrstraps
nboot = 200
```

# Clonal abundance

For plotting the clonal abundance, the clones were ordered by size from bigger clones to smaller clones (x-axis, Rank).
The Abundance of each clone was represented as the percentage of unique sequences in the clone, with respect to the total
number of unique sequences in that subject (By Patient) or in the B-cell or T-cell sample (By Cell Population).

To correct for the different number of sequences in each of the samples, the Bootstrapping technique was employed,
in which `r nboot` random bootstrap samples were taken, with size the number of sequences in the sample with less sequences (N).
The solid line shows the mean Abundance of the bootstrap samples, whereas the transparent area shows the full Abundance
range of the bootstrap samples.

All clonal abundance plots and tables with abundance values can be found under `repertoire_analysis/Abundance`.

## Clonal abundance per subject
```{r clonal_abundance, echo=FALSE}
abund <- estimateAbundance(df_all, group = "subj_locus", ci=0.95, nboot=nboot)
abund@abundance$sample_id <- sapply(abund@abundance$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[1])
abund@abundance$subject_id <- sapply(abund@abundance$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[2])
abund@abundance$locus <- sapply(abund@abundance$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[3])


abund_main <- paste0("Clonal abundance (N=", abund@n[1], ")")

p_ca <- ggplot(abund@abundance, aes(x = rank, y = p)) +
    geom_ribbon(aes(ymin = lower,
                    ymax = upper, fill = sample_id), alpha = 0.4) +
    geom_line(aes(color = sample_id)) +
    ggtitle(abund_main) +
    xlab("Rank") + ylab("Abundance") +
    scale_x_log10(limits = NULL,
                breaks = scales::trans_breaks("log10", function(x) 10^x),
                labels = scales::trans_format("log10", scales::math_format(10^.x))) +
    scale_y_continuous(labels = scales::percent) #+
    #facet_grid(cols = vars(locus), rows = vars(subject_id), scales="free", drop = T)
ggplotly(p_ca)
```

```{r plot_abundance, include = FALSE}
ggsave(plot=p_ca, filename = paste0(abundance_dir,"/Clonal_abundance_subject.pdf"), device="pdf", width = 25, height = 10, units="cm")
ggsave(plot=p_ca, filename = paste0(abundance_dir,"/Clonal_abundance_subject.png"), device="png", width = 25, height = 10, units="cm")
write.table(abund@abundance, file = paste0(abundance_dir, "/Clonal_abundance_data_subject.tsv"), sep="\t", quote = F, row.names = F)
```

# Clonal diversity

The clonal diversity $D$ of the repertoire was calculated according to the general formula of Hill Diversity
numbers:

$$
\begin{aligned}
    ^{q}D = \left( \sum_{i=1}^Rp_i^q \right)^{1/(1-q)}
\end{aligned}
$$

where:

* $p_i$ is the proportion of unique sequences belonging to clone $i$.
* $q$ are the values of the different diversity numbers.
* $R$ is the Richness, the number of different clones in the sample.

At $q=1$ the function is undefined and the limit to zero equals the exponential of the Shannon Entropy:

$$
\begin{aligned}
    ^{1}D = exp \left(  \sum_{i=1}^Rp_i ln(p_i)  \right)
\end{aligned}
$$

The intuition about the different Hill Diversity values is the following:

* At $q=0$ the diversity index equals the number of clones in the sample.
* At $q=1$ the diversity index is the geometric mean of the clones in the sample,
weighted by their proportion in the sample.
* At $q>1$ more weight is given to the clones with higher proportions in the sample.

All clonal diversity plots and tables with diversity values can be found under `repertoire_analysis/Diversity`.
To correct for the different number of sequences in each of the samples, the Bootstrapping technique was employed,
in which `r nboot` random bootstrap samples were taken, with size the number of sequences in the sample with less sequences (N).
The solid line shows the mean Diversity of the bootstrap samples, whereas the transparent area shows the full Diversity
range of the bootstrap samples.


## Clonal diversity per subject
```{r clonal_diversity, echo = FALSE}
sample_div <- alphaDiversity(abund, group="subj_locus", min_q=0, max_q=4, step_q=0.05,
                            ci=0.95, nboot=nboot)
sample_main <- paste0("Sample diversity (N=", sample_div@n[1], ")")

sample_div@diversity$sample_id <- sapply(sample_div@diversity$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[1])
sample_div@diversity$subject_id <- sapply(sample_div@diversity$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[2])
sample_div@diversity$locus <- sapply(sample_div@diversity$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[3])

div_p <- ggplot(sample_div@diversity, aes(x = q, y = d, group=sample_id)) +
    geom_ribbon(aes(ymin = d_lower,
                    ymax = d_upper,
                    fill = sample_id), alpha = 0.4) +
    geom_line(aes(color = sample_id)) +
    xlab("q") + ylab("Diversity(q)") +
    ggtitle(sample_main) #+
    #facet_grid(cols=vars(locus), rows=vars(subject_id))
ggplotly(div_p)
```
```{r plot_diversity, include = FALSE}
ggsave(plot=div_p, filename=paste0(diversity_dir,"/Diversity_patient_grid.png"), device="png", width = 25, height = 10, units="cm")
ggsave(plot=div_p, filename=paste0(diversity_dir,"/Diversity_patient_grid.pdf"), device="pdf", width = 25, height = 10, units="cm")
write.table(sample_div@diversity, file = paste0(diversity_dir, "/Clonal_diversity_data_subject.tsv"), sep="\t", quote = F, row.names = F)
```

# V gene usage

## V gene family usage

The V gene usage (in percentage) in each of the samples is represented below. All plots and tables can be found [here](./repertoire_comparison/V_family/).

Gene family usage is normalized by the number of clones.

```{r gene_family, echo=F}
family <- countGenes(df_all, gene="v_call", groups="subj_locus", mode="family", clone="clone_id")
family$sample_id <- sapply(family$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[1])
family$subject_id <- sapply(family$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[2])
family$locus <- sapply(family$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[3])

g2 <- ggplot(family, aes(x=gene, y=clone_freq, fill=sample_id, group=sample_id)) +
    scale_color_brewer(palette="Set1") +
    geom_bar(position = "dodge", stat="identity") +
    ggtitle("V Gene Family Usage") +
    theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
    ylab("Frequency") +
    xlab("") +
    #facet_grid(cols=vars(locus), rows=vars(subject_id)) +
    theme(legend.position = "right")
ggplotly(g2)

ggsave(filename = paste0(vfamily_dir, "/V_Family_distribution_patient.svg"), plot = g2, width = 18, height = 15, units = "cm")
ggsave(filename = paste0(vfamily_dir, "/V_Family_distribution_patient.png"), plot = g2, width = 18, height = 15, units = "cm")

write.table(family, file = paste0(vfamily_dir, "/V_family_distribution_data.tsv"), sep = "\t", quote = F, row.names = F)
```

## V gene usage {.tabset .tabset-fade .tabset-pills}

The V gene usage (in percentage) in each of the samples is represented below.
All plots and tables can be found [here](./repertoire_comparison/V_family/).

### By clones {-}

```{r gene_clonal, echo=F}
family <- countGenes(df_all, gene="v_call", groups="subj_locus", mode="gene", clone = "clone_id")
family$sample_id <- sapply(family$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[1])
family$subject_id <- sapply(family$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[2])
family$locus <- sapply(family$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[3])

g2 <- ggplot(family, aes(x=gene, y=clone_freq, fill=sample_id, group=sample_id)) +
    scale_color_brewer(palette="Set1") +
    geom_bar(position = "dodge", stat="identity") +
    ggtitle("V Gene Family Usage") +
    theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
    ylab("Frequency") +
    xlab("") #+
    #facet_wrap(vars(subject_id,locus), scales="free_x", ncol = 1) +
    #theme(legend.position = "none")
ggplotly(g2)
ggsave(filename = paste0(vfamily_dir, "/V_gene_distribution_by_clone_patient.svg"), plot = g2, width = 20, height = 40, units = "cm")
ggsave(filename = paste0(vfamily_dir, "/V_gene_distribution_by_clone_patient.png"), plot = g2, width = 20, height = 40, units = "cm")

write.table(family, file = paste0(vfamily_dir, "/V_gene_distribution_by_clone_data.tsv"), sep = "\t", quote = F, row.names = F)
```

### By sequences {-}

```{r gene_seq, echo=F}
family <- countGenes(df_all, gene="v_call", groups="subj_locus", mode="gene")
family$sample_id <- sapply(family$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[1])
family$subject_id <- sapply(family$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[2])
family$locus <- sapply(family$subj_locus, function(x) unlist(strsplit(as.character(x), "_"))[3])

g2 <- ggplot(family, aes(x=gene, y=seq_freq, fill=sample_id, group=sample_id)) +
    scale_color_brewer(palette="Set1") +
    geom_bar(position = "dodge", stat="identity") +
    ggtitle("V Gene Usage") +
    theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1)) +
    ylab("Frequency") +
    xlab("") #+
    #facet_wrap(vars(subject_id,locus), scales="free_x", ncol = 1) +
    #theme(legend.position = "none")
ggplotly(g2)
ggsave(filename = paste0(vfamily_dir, "/V_gene_distribution_by_sequence_patient.svg"), plot = g2, width = 20, height = 40, units = "cm")
ggsave(filename = paste0(vfamily_dir, "/V_gene_distribution_by_sequence_patient.png"), plot = g2, width = 20, height = 40, units = "cm")

write.table(family, file = paste0(vfamily_dir, "/V_gene_distribution_by_sequence_data.tsv"), sep = "\t", quote = F, row.names = F)
```

# Citations

If you use nf-core/airrflow for your analysis, please cite it using the following DOI: [10.5281/zenodo.3607408](https://doi.org/10.5281/zenodo.3607408)

Please also cite the `nf-core` and `Nextflow` publications:

- [nf-core](https://pubmed.ncbi.nlm.nih.gov/32055031/)

  > Ewels PA, Peltzer A, Fillinger S, Patel H, Alneberg J, Wilm A, Garcia MU, Di Tommaso P, Nahnsen S. The nf-core framework for community-curated bioinformatics pipelines. Nat Biotechnol. 2020 Mar;38(3):276-278. doi: 10.1038/s41587-020-0439-x. PubMed PMID: 32055031.

- [Nextflow](https://pubmed.ncbi.nlm.nih.gov/28398311/)

  > Di Tommaso P, Chatzou M, Floden EW, Barja PP, Palumbo E, Notredame C. Nextflow enables reproducible computational workflows. Nat Biotechnol. 2017 Apr 11;35(4):316-319. doi: 10.1038/nbt.3820. PubMed PMID: 28398311.

In addition, citations for the tools and data used in this pipeline are as follows:

- [pRESTO](https://doi.org/10.1093/bioinformatics/btu138)

  > Vander Heiden, J. A., Yaari, G., Uduman, M., Stern, J. N. H., O’Connor, K. C., Hafler, D. A., … Kleinstein, S. H. (2014). pRESTO: a toolkit for processing high-throughput sequencing raw reads of lymphocyte receptor repertoires. Bioinformatics, 30(13), 1930–1932.

- [SHazaM, Change-O](https://doi.org/10.1093/bioinformatics/btv359)

  > Gupta, N. T., Vander Heiden, J. A., Uduman, M., Gadala-Maria, D., Yaari, G., & Kleinstein, S. H. (2015). Change-O: a toolkit for analyzing large-scale B cell immunoglobulin repertoire sequencing data: Table 1. Bioinformatics, 31(20), 3356–3358.

- [Alakazam](https://doi.org/10.1126/scitranslmed.3008879)

  > Stern, J. N. H., Yaari, G., Vander Heiden, J. A., Church, G., Donahue, W. F., Hintzen, R. Q., … O’Connor, K. C. (2014). B cells populating the multiple sclerosis brain mature in the draining cervical lymph nodes. Science Translational Medicine, 6(248).

- [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/)

- [MultiQC](https://pubmed.ncbi.nlm.nih.gov/27312411/)

  > Ewels P, Magnusson M, Lundin S, Käller M. MultiQC: summarize analysis results for multiple tools and samples in a single report. Bioinformatics. 2016 Oct 1;32(19):3047-8. doi: 10.1093/bioinformatics/btw354. Epub 2016 Jun 16. PubMed PMID: 27312411; PubMed Central PMCID: PMC5039924.

## Software packaging/containerisation tools

- [Anaconda](https://anaconda.com)

  > Anaconda Software Distribution. Computer software. Vers. 2-2.4.0. Anaconda, Nov. 2016. Web.

- [Bioconda](https://pubmed.ncbi.nlm.nih.gov/29967506/)

  > Grüning B, Dale R, Sjödin A, Chapman BA, Rowe J, Tomkins-Tinch CH, Valieris R, Köster J; Bioconda Team. Bioconda: sustainable and comprehensive software distribution for the life sciences. Nat Methods. 2018 Jul;15(7):475-476. doi: 10.1038/s41592-018-0046-7. PubMed PMID: 29967506.

- [BioContainers](https://pubmed.ncbi.nlm.nih.gov/28379341/)

  > da Veiga Leprevost F, Grüning B, Aflitos SA, Röst HL, Uszkoreit J, Barsnes H, Vaudel M, Moreno P, Gatto L, Weber J, Bai M, Jimenez RC, Sachsenberg T, Pfeuffer J, Alvarez RV, Griss J, Nesvizhskii AI, Perez-Riverol Y. BioContainers: an open-source and community-driven framework for software standardization. Bioinformatics. 2017 Aug 15;33(16):2580-2582. doi: 10.1093/bioinformatics/btx192. PubMed PMID: 28379341; PubMed Central PMCID: PMC5870671.

- [Docker](https://dl.acm.org/doi/10.5555/2600239.2600241)

- [Singularity](https://pubmed.ncbi.nlm.nih.gov/28494014/)

  > Kurtzer GM, Sochat V, Bauer MW. Singularity: Scientific containers for mobility of compute. PLoS One. 2017 May 11;12(5):e0177459. doi: 10.1371/journal.pone.0177459. eCollection 2017. PubMed PMID: 28494014; PubMed Central PMCID: PMC5426675.

