---
output:
  html_document:
    toc: true                               # table of contents
    toc_float: true                         # float the table of contents to the left of the main document content
    toc_depth: 3                            # header levels 1,2,3
    theme: default
    number_sections: false                  # add section numbering to headers
    df_print: paged                         # tables are printed as an html table with support for pagination over rows and columns
    css: ./nf-core_style.css
    highlight: pygments
    pdf_document: true
  html_notebook:
    toc: yes
---


---
title: "Airrflow results summary report"
subtitle: "Interactive summary report produced by the nf-core/airrflow pipeline"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---


```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(tidyr)
library(alakazam)
library(shazam)
library(stringr)
library(plotly)
library(airr)
library(DT)
library(scales)

ggplot_theme <- theme_bw(base_family = "ArialMT") +
            theme(panel.grid.major = element_blank(),
            panel.border = element_rect(colour = "black", fill=NA, size=0.5),
            panel.grid.minor = element_blank(), text = element_text(family="ArialMT"))
theme_set(ggplot_theme)
knitr::opts_chunk$set(echo = FALSE)

# x-tick info text
step_help <- list(
  # Sequence assembly steps
  "Sequences"           = "Number of read pairs after read quality control and adapter trimming.",
  "Filtered_quality"    = "Number of reads with base quality above --filterseq_q.",
  "Filtered_quality_R1" = "Number of R1 reads with base quality above --filterseq_q.",
  "Filtered_quality_R2" = "Number of R2 reads with base quality above --filterseq_q.",
  "Mask_primers_R1"     = "Number of R1 reads with matching primer in the primer file.",
  "Mask_primers_R2"     = "Number of R2 reads with matching primer in the primer file.",
  "Paired"              = "Number of read pairs for which R1 and R2 passed the filtering steps.",
  "Build_consensus"     = "Number of consensus sequences. Multiple reads sharing a barcode are collapsed into one consensus sequence. When the UMI diversity is insufficient sequences are optionally clustered according to similarity before a consensus sequence is built. A drop in the number of sequences is expected in this step. Dependent on --primer_consensus,--buildconsensus_maxerror and --buildconsensus_maxgap.",
  "Assemble_pairs"      = "Number of successful sequence assemblies.",
  "Unique"              = "Number of unique assemblies after collapsing exact duplicates. Identical sequences are merged and their read counts summed.",
  "Representative_2"    = "Number of unique assemblies supported by at least two reads. Assemblies seen only once are discarded.",
  # V(D)J gene assignment and QC steps
  "AssignGenes-igblast" = "Number of assemblies for which gene segments are assigned with igblast.",
  "MakeDB-igblast"      = "Number of assemblies for which igblast returned functionality information, V and J gene assignment, and a junction region.",
  "FilterQuality"       = "Number of assemblies for which the provided locus matches the V gene assignment and the igblast alignment had a minimum of 200 informative positions and a maximum of 10% N nucleotides.",
  "ParseDb-split"       = "Number of assemblies identified as functional rearrangements by IgBLAST.",
  "FilterJunctionMod3"  = "Number of assemblies with a junction length divisible by 3.",
  "AddMetadata"         = "Number of assemblies with annotated metadata.",
  "SingleCellQC"        = "Number of single-cell assemblies after removing contaminants and assemblies having no or multiple heavy chains.",
  "RemoveChimeric"      = "Number of assemblies that are not chimeric.",
  "CollapseDuplicates"  = "Number of assemblies after deduplication.",
  "CreateGermlines"     = "Number of assemblies for which a germline sequence could be reconstructed from the assigned V, D and J genes.",
  "ClonePass"           = "Number of assemblies assigned to a clonal group."
)

# add info text to x-ticks
tick_help_js <- '
function(el, x, data) {

  // create div
  var panel = document.createElement("div");
  panel.className = "airrflow-help-panel";
  panel.style.display = "none";
  el.insertAdjacentElement("afterend", panel);

  function hide() {
    panel.style.display = "none";
    panel.airrflowTick = null;
  }

  function helpTicks() {
    var ticks = [];
    el.querySelectorAll(".xtick text").forEach(function(node) {
      var key = (node.textContent || "").trim();
      if (Object.prototype.hasOwnProperty.call(data, key)) {
        node.classList.add("airrflow-help-tick");
        ticks.push({ node: node, key: key });
      }
    });
    return ticks;
  }

  helpTicks();
  new MutationObserver(helpTicks).observe(el, { childList: true, subtree: true });

  el.addEventListener("click", function(ev) {
    var hit = null;
    helpTicks().forEach(function(tick) {
      if (tick.node === ev.target || tick.node.contains(ev.target)) {
        hit = tick;
      }
    });

    if (!hit) { hide(); return; }

    // Create div content
    panel.innerHTML = "";
    var title = document.createElement("strong");
    title.textContent = hit.key;
    var body = document.createElement("div");
    body.textContent = data[hit.key];
    panel.appendChild(title);
    panel.appendChild(body);
    panel.airrflowTick = hit.node;
    panel.style.display = "block";
  }, true);
}'

# Show only the samples selected in the metadata table
sample_filter_js <- '
function(el, x, col) {
  var dt = window.jQuery && jQuery("#metadata-table table").DataTable();
  if (!dt) return;
  function apply() {
    var keep = dt.rows({selected: true}).data().toArray().map(function(r) { return r[col]; });
    Plotly.restyle(el, {visible: el.data.map(function(t) {
      return keep.length === 0 || keep.indexOf(t.name) > -1;
    })});
  }
  dt.on("select deselect", apply);
  setTimeout(apply, 0);
}'

# Setting data table and creating output folders
outdir <- "repertoire_comparison"
dir.create(outdir)
seq_dir <- paste(outdir, "Sequence_numbers_summary", sep="/")
dir.create(seq_dir)

# Read data from the work directory
datadir <- "."

```

A full description of the pipeline results can be found on the Output section of the [nf-core/airrflow website](https://nf-co.re/airrflow).

# Metadata

The table below shows the metadata provided in the input sample sheet for this run.

```{r sample_metadata, echo=FALSE, warning=FALSE}
samplesheet <- read.table("./samplesheet.valid.tsv", header = TRUE, sep = "\t", check.names = FALSE)
sample_col <<- which(names(samplesheet) == "sample_id") - 1
DT::datatable(samplesheet, rownames = FALSE, elementId = "metadata-table",
              extensions = "Select", selection = "none",
              options = list(select = list(style = "multi"), scrollX = TRUE,
                              scrollY = "400px", paging = FALSE, scrollCollapse = TRUE))
```

# Seqeuencing depth
Reads per UMI plotted for each of the samples.
```{r reads_per_UMI_plot, echo=FALSE, warning=FALSE, results='asis'}
tryCatch( {
    tab_seq  <- read.table("reads_per_UMI.csv", header=TRUE, sep=",")
    tab_seq[is.na(tab_seq)] <- 0
    plot_table <- tidyr::pivot_longer(tab_seq,
                                      cols=-X,
                                      names_to = "sample_id",
                                      values_to ="count")

    x_range <- c(floor(min(plot_table$X)), ceiling(max(plot_table$X)))
    y_max <- ceiling(max(log10(plot_table$count)))
    UMI_plot <- ggplot(data=plot_table,
                        aes(x=X, y=count, group=sample_id)) +
                  geom_line(aes(colour=sample_id)) +
                  geom_point(aes(colour=sample_id)) +
                  xlab('Reads per UMI') + ylab('UMI count') +
                  scale_x_reverse(breaks=seq(x_range[1], x_range[2], by=ceiling(x_range[2]/10))) +
                  scale_y_continuous(
                    trans=scales::pseudo_log_trans(sigma=1, base=10),
                    breaks=c(0,10^seq(1, y_max,length.out=max(3,y_max)))
                  )

    p <- htmlwidgets::onRender(ggplotly(UMI_plot), sample_filter_js, data = sample_col)
  p

  },
  error=function(e){message("No reads per UMI information available.")}
)
```

# Number of sequences

## Sequence assembly steps

When provided with raw read data, the pipeline performs read quality control and sequence assembly.
The line plot below shows the number or reads and assemblies for each of the samples after each of the steps performed.
The full table can be found under [Table_sequences_assembly](repertoire_comparison/Sequence_numbers_summary/Table_sequences_assembly.tsv).
Click the plot labels for a more detailed description.

```{r seq_numbers_plot, echo=FALSE, warning=FALSE, results='asis'}
tryCatch( {
    tab_seqs <- read.table("./Table_sequences.tsv", header=TRUE, sep="\t", check.names = FALSE)
    write.table(tab_seqs, file=paste0(seq_dir,"/Table_sequences_assembly.tsv"), sep="\t", quote=F, row.names=F)

    plot_table <- tidyr::pivot_longer(tab_seqs,
                                      cols=Sequences:Representative_2,
                                      names_to = "steps",
                                      values_to ="count")
    firstcol = which(colnames(tab_seqs) == "Sequences")
    lastcol = which(colnames(tab_seqs) == "Representative_2")
    plot_table$steps <- factor(plot_table$steps, levels=colnames(tab_seqs)[firstcol:lastcol])

    y_min <- floor(scales::pseudo_log_trans(sigma=1,base=10)$transform(min(plot_table$count)))
    y_max <- max(ceiling(max(log10(plot_table$count))),y_min+2)
    seqs_plot <- ggplot(data=plot_table,
                        aes(x=steps, y=count, group=sample_id)) +
                    geom_line(aes(colour=sample_id)) +
                    geom_point(aes(colour=sample_id)) +
                    scale_y_continuous(
                        trans=scales::pseudo_log_trans(sigma=1,base=10),
                        breaks=10^seq(y_min, y_max,by=1),
                        limits=c(min(10^y_min,min(plot_table$count)),10^y_max)
                        ) +
                    xlab("") + ylab("Sequence count") +
                    ggtitle("Number of sequences after each of the sequence assembly steps") +
                    theme(axis.text.x= element_text(angle = 45))

    p <- htmlwidgets::onRender(ggplotly(seqs_plot), tick_help_js, data = step_help)
    p <- htmlwidgets::onRender(p, sample_filter_js, data = sample_col)
    p
  },
  error=function(e){message("No sequence numbers are available if starting with assembled reads.")}
)
```

## V(D)J gene assignment and QC

When provided with assembled sequences or after performing sequence assembly, the pipeline performs V(D)J gene annotation and quality control of alignments used for gene assignment.
The line plot below displays the number of sequences for each of the samples after each of the downstream filtering and clonal analysis steps.
The full table can be found under [Table_sequences_assembled](repertoire_comparison/Sequence_numbers_summary/Table_sequences_assembled.tsv).
Click the plot labels for a more detailed description.

```{r seq_assembled, echo=FALSE, warning=FALSE, results='asis'}
tab_seqs_assembled <- read.table("./Table_sequences_assembled.tsv", header=TRUE, sep="\t", check.names = FALSE)

# Splitting on last underscore for sample_id and if there is no underscore split on last dot
if (is.null(tab_seqs_assembled$sample_id)) {
  tab_seqs_assembled$sample_id <- sapply(tab_seqs_assembled$input, function(x) {
    if (grepl("_", x)) {
      unlist(strsplit(as.character(x), "_(?=[^_]+$)", perl=TRUE))[1]
    } else if (grepl("\\.", x)) {
      unlist(strsplit(as.character(x), "\\.(?=[^\\.]+$)", perl=TRUE))[1]
    } else {
      x
    }
  })
}

dat <- tab_seqs_assembled %>%
        filter( !grepl("-fail.tsv", output) ) %>%
        filter( !grepl("productive-F.tsv", output) ) %>%
        tidyr::pivot_wider(id_cols=sample_id,
                            names_from=task,
                            values_from=output_size)
dat <- dat %>% dplyr::select(any_of(c("sample_id","ConvertDb-fasta", "AssignGenes-igblast", "MakeDB-igblast", "FilterQuality",
                                "ParseDb-split", "FilterJunctionMod3","AddMetadata","SingleCellQC","CreateGermlines",
                                "RemoveChimeric","CollapseDuplicates","ClonePass"))) %>% dplyr::arrange(sample_id)

dat <- apply(dat,2,as.character)
write.table(dat, file=paste0(seq_dir,"/Table_sequences_assembled.tsv"), sep="\t", quote=F, row.names=F)
```


```{r assembled_seq_numbers_plot, echo=FALSE, warning=FALSE, results='asis'}

tab_seqs_assembled <- tab_seqs_assembled %>%
        filter( !grepl("-fail.tsv", output) ) %>%
        filter( !grepl("productive-F.tsv", output) ) %>%
        dplyr::filter( task %in% c("sample_id","AssignGenes-igblast", "MakeDB-igblast", "FilterQuality",
                                "ParseDb-split", "FilterJunctionMod3","AddMetadata","SingleCellQC","CreateGermlines",
                                "RemoveChimeric","CollapseDuplicates","ClonePass"))

tab_seqs_assembled$task <- factor(tab_seqs_assembled$task, levels=c("AssignGenes-igblast", "MakeDB-igblast", "FilterQuality",
                                                        "ParseDb-split", "FilterJunctionMod3", "AddMetadata", "SingleCellQC",
                                                        "CreateGermlines", "RemoveChimeric", "CollapseDuplicates",
                                                        "ClonePass"))

tab_seqs_assembled$output_size <- as.numeric(tab_seqs_assembled$output_size)
# Remove rows with NA or zero output_size
tab_seqs_assembled <- tab_seqs_assembled %>%
  filter(!is.na(output_size))

y_min <- floor(scales::pseudo_log_trans(sigma=1,base=10)$transform(min(tab_seqs_assembled$output_size)))
y_max <- max(ceiling(max(log10(tab_seqs_assembled$output_size))),y_min+2)
seqs_plot_assembled <- ggplot(data=tab_seqs_assembled,
                    aes(x=task, y=output_size, group=sample_id)) +
              geom_line(aes(colour=sample_id)) +
              geom_point(aes(colour=sample_id)) +
              scale_y_continuous(
                    trans=scales::pseudo_log_trans(sigma=1,base=10),
                    breaks=10^seq(y_min, y_max,by=1),
                    limits=c(min(10^y_min,min(tab_seqs_assembled$output_size)),10^y_max)
              ) +
              xlab("") + ylab("Sequence count") +
              ggtitle("Number of sequences after each of the downstream steps") +
              theme(axis.text.x = element_text(angle = 45)
              )

p <- htmlwidgets::onRender(ggplotly(seqs_plot_assembled), tick_help_js, data = step_help)
p <- htmlwidgets::onRender(p, sample_filter_js, data = sample_col)
p
```

# Clonal analysis

The pipeline performs clonal assignment using hierarchical clustering.

## Clustering threshold identification
When `--clonal_threshold` is set to 'auto', a hamming distance threshold for cluster assignment is estimated.
A report showing the estimated thresholds and nearest-neighbor distributions of all subjects can be found at [clonal_analysis/find_threshold/all_reps_dist_report/index.html](clonal_analysis/find_threshold/all_reps_dist_report/index.html).

## Clonal assignment
Reports summarizing the clonal assignment for all subjects can be found at [clonal_analysis/clonal_assignment/\<subjectID\>_clone_report/index.html](clonal_analysis/clonal_assignment/).

## AIRR repertoire_analysis
A report summarizing the repertoire properties for all the samples can be found at [clonal_analysis/repertoire_analysis/repertoire_analysis_report/index.html](clonal_analysis/repertoire_analysis/repertoire_analysis_report/index.html).

## Clone lineage trees
Reports summarizing the lineages trees of each of the specified clonal groups can be found at [clonal_analysis/dowser_lineages/\<sampleID\>_dowser_report/index.html](clonal_analysis/dowser_lineages/).



# Citations

If you use nf-core/airrflow for your analysis, please cite our paper:

> Gisela Gabernet, Susanna Marquez, Robert Bjornson, Alexander Peltzer, Hailong Meng, Edel Aron, Noah Y. Lee, Cole G. Jensen, David Ladd, Mark Polster, Friederike Hanssen, Simon Heumos, nf-core community, Gur Yaari, Markus C. Kowarik, Sven Nahnsen, Steven H. Kleinstein. nf-core/airrflow: an adaptive immune receptor repertoire analysis workflow employing the Immcantation framework. (2024) PLOS Computational Biology, 20(7), e1012265. doi: [https://doi.org/10.1371/journal.pcbi.1012265](https://doi.org/10.1371/journal.pcbi.1012265). Pubmed PMID: 39058741.

In addition, citations for the tools and data used in this pipeline are as follows:

- [pRESTO](https://doi.org/10.1093/bioinformatics/btu138)

  > Vander Heiden, J. A., Yaari, G., Uduman, M., Stern, J. N. H., O’Connor, K. C., Hafler, D. A., … Kleinstein, S. H. (2014). pRESTO: a toolkit for processing high-throughput sequencing raw reads of lymphocyte receptor repertoires. Bioinformatics, 30(13), 1930–1932.

- [SHazaM, Change-O](https://doi.org/10.1093/bioinformatics/btv359)

  > Gupta, N. T., Vander Heiden, J. A., Uduman, M., Gadala-Maria, D., Yaari, G., & Kleinstein, S. H. (2015). Change-O: a toolkit for analyzing large-scale B cell immunoglobulin repertoire sequencing data: Table 1. Bioinformatics, 31(20), 3356–3358.

- [IgBLAST](https://doi.org/10.1093/nar/gkt382)

  > Ye, J., Ma, N., Madden, T. L., & Ostell, J. M. (2013). IgBLAST: An immunoglobulin variable domain sequence analysis tool. Nucleic Acids Research, 41(Web Server issue), W34.

- [Alakazam](https://doi.org/10.1126/scitranslmed.3008879)

  > Stern, J. N. H., Yaari, G., Vander Heiden, J. A., Church, G., Donahue, W. F., Hintzen, R. Q., … O’Connor, K. C. (2014). B cells populating the multiple sclerosis brain mature in the draining cervical lymph nodes. Science Translational Medicine, 6(248), 248ra107.

- [SCOPer](https://doi.org/10.1093/bioinformatics/bty235)

  > Nouri N, Kleinstein S (2018). “A spectral clustering-based method for identifying clones from high-throughput B cell repertoire sequencing data.” Bioinformatics, i341-i349.

  > Nouri N, Kleinstein S (2020). “Somatic hypermutation analysis for improved identification of B cell clonal families from next-generation sequencing data.” PLOS Computational Biology, 16(6), e1007977.

  > Gupta N, Adams K, Briggs A, Timberlake S, Vigneault F, Kleinstein S (2017). “Hierarchical clustering can identify B cell clones with high confidence in Ig repertoire sequencing data.” The Journal of Immunology, 2489-2499.

- [Dowser](https://doi.org/10.1371/journal.pcbi.1009885)

  > Hoehn K, Pybus O, Kleinstein S (2022). “Phylogenetic analysis of migration, differentiation, and class switching in B cells.” PLoS Computational Biology.

- [IgPhyML](https://www.pnas.org/doi/10.1073/pnas.1906020116)

  > Hoehn K, Van der Heiden J, Zhou J, Lunter G, Pybus O, Kleinstein S (2019). “Repertoire-wide phylogenetic models of B cell molecular evolution reveal evolutionary signatures of aging and vaccination.” PNAS.

- [RAxML](10.1093/bioinformatics/btu033)

  > Stamatakis A. (2014) RAxML version 8: a tool for phylogenetic analysis and post-analysis of large phylogenies. Bioinformatics, 30(9): 1312-1313.

- [TIgGER](https://doi.org/10.1073/pnas.1417683112)

  > Gadala-maria, D., Yaari, G., Uduman, M., & Kleinstein, S. H. (2015). Automated analysis of high-throughput B-cell sequencing data reveals a high frequency of novel immunoglobulin V gene segment alleles. Proceedings of the National Academy of Sciences, 112(8), 1–9.

- [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/)

- [Fastp](https://doi.org/10.1093/bioinformatics/bty560)

  > Shifu Chen, Yanqing Zhou, Yaru Chen, Jia Gu, fastp: an ultra-fast all-in-one FASTQ preprocessor, Bioinformatics. 2018 Sept 1; 34(17):i884–i890.

- [MultiQC](https://pubmed.ncbi.nlm.nih.gov/27312411/)

  > Ewels P, Magnusson M, Lundin S, Käller M. MultiQC: summarize analysis results for multiple tools and samples in a single report. Bioinformatics. 2016 Oct 1;32(19):3047-8. doi: 10.1093/bioinformatics/btw354. Epub 2016 Jun 16. PubMed PMID: 27312411; PubMed Central PMCID: PMC5039924.

## Software packaging/containerisation tools

- [Anaconda](https://anaconda.com)

  > Anaconda Software Distribution. Computer software. Vers. 2-2.4.0. Anaconda, Nov. 2016. Web.

- [Bioconda](https://pubmed.ncbi.nlm.nih.gov/29967506/)

  > Grüning B, Dale R, Sjödin A, Chapman BA, Rowe J, Tomkins-Tinch CH, Valieris R, Köster J; Bioconda Team. Bioconda: sustainable and comprehensive software distribution for the life sciences. Nat Methods. 2018 Jul;15(7):475-476. doi: 10.1038/s41592-018-0046-7. PubMed PMID: 29967506.

- [BioContainers](https://pubmed.ncbi.nlm.nih.gov/28379341/)

  > da Veiga Leprevost F, Grüning B, Aflitos SA, Röst HL, Uszkoreit J, Barsnes H, Vaudel M, Moreno P, Gatto L, Weber J, Bai M, Jimenez RC, Sachsenberg T, Pfeuffer J, Alvarez RV, Griss J, Nesvizhskii AI, Perez-Riverol Y. BioContainers: an open-source and community-driven framework for software standardization. Bioinformatics. 2017 Aug 15;33(16):2580-2582. doi: 10.1093/bioinformatics/btx192. PubMed PMID: 28379341; PubMed Central PMCID: PMC5870671.

- [Docker](https://dl.acm.org/doi/10.5555/2600239.2600241)

- [Singularity](https://pubmed.ncbi.nlm.nih.gov/28494014/)

  > Kurtzer GM, Sochat V, Bauer MW. Singularity: Scientific containers for mobility of compute. PLoS One. 2017 May 11;12(5):e0177459. doi: 10.1371/journal.pone.0177459. eCollection 2017. PubMed PMID: 28494014; PubMed Central PMCID: PMC5426675.
